A Large-scale Study of Polymorphism in R

### 2019-04-04 notebooks/data_builder.Rmd
new pipeline is happening, dealing with the data. final numbers and adjusted discussion will have to be done tomorrow

>>>>

```{r}
# collapse
# make sure the right function is uncommented in Java

psp_col <- read_csv("../data/partial_L1_collapsed.csv.gz")
psp_L0_col <- read_csv("../data/partial_L0_collapsed.csv.gz")
```

```{r}
# arg counts
# again, Java
L0_arg_sig_c <- read_csv("../data/partial_L0_arg_sig_counts.csv")
L1_arg_sig_c <- read_csv("../data/partial_L1_arg_sig_counts.csv")

# sort
L0_arg_sig_c[order(-L0_arg_sig_c$count),] -> L0_arg_sig_c
L1_arg_sig_c[order(-L1_arg_sig_c$count),] -> L1_arg_sig_c

# rm []
L0_arg_sig_c[1,"count"] <- 0
L1_arg_sig_c[1,"count"] <- 0

# re sort
L0_arg_sig_c[order(-L0_arg_sig_c$count),] -> L0_arg_sig_c
L1_arg_sig_c[order(-L1_arg_sig_c$count),] -> L1_arg_sig_c

# percentages
L0_arg_sig_c$perc_tot <- round(L0_arg_sig_c$count / sum(L0_arg_sig_c$count) * 100, 2)
L1_arg_sig_c$perc_tot <- round(L1_arg_sig_c$count / sum(L1_arg_sig_c$count) * 100, 2)

# collapse the obvious signatures
L0_arg_sig_c <- new_df_count_collape_arg_sigs(L0_arg_sig_c, "L0")
L1_arg_sig_c <- new_df_count_collape_arg_sigs(L1_arg_sig_c, "L1")

# poly arg counts only
L0_poly_sig <- L0_arg_sig_c[L0_arg_sig_c$arg_sig %>% strsplit(split=",") %>% lapply(function(x) length(x) != 1) %>% unlist, ]
L1_poly_sig <- L1_arg_sig_c[L1_arg_sig_c$arg_sig %>% strsplit(split=",") %>% lapply(function(x) length(x) != 1) %>% unlist, ]

# display
L0_poly_sig[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
L1_poly_sig[1:10,] %>% ready_print_count_fun %>% xtable %>% print(include.rownames=F)
```

```{r}
# all pckg (w/ dups)
pf_L0 <- psp_L0 %>% mutate(pkgfun = paste(pkg, fun, sep="::")) %>% select(pkgfun)
pf_L1 <- psp %>% mutate(pkgfun = paste(pkg, fun, sep="::")) %>% select(pkgfun)

# tot fun counts
pf_L0 %>% unique %>% nrow -> tot_L0_funs
pf_L1 %>% unique %>% nrow -> tot_L1_funs

# poly fun counts
pf_L0[duplicated(pf_L0),] %>% unique -> poly_L0_funs
pf_L1[duplicated(pf_L1),] %>% unique -> poly_L1_funs

poly_L0_funs_c <- poly_L0_funs %>% nrow
poly_L1_funs_c <- poly_L1_funs %>% nrow

# mono fun count (sanity check)
setdiff(pf_L0 %>% unique, poly_L0_funs) %>% nrow -> mono_L0_funs_c
setdiff(pf_L1 %>% unique, poly_L1_funs) %>% nrow -> mono_L1_funs_c

data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L0_funs_c/1000, tot_L0_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L0_funs_c/tot_L0_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
```

```{r}
# arg poly count

# L0
data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=c(L0_arg_sig_c$count %>% sum - L0_poly_sig$count %>% sum, L0_arg_sig_c$count %>% sum), Percentage=c(paste0(round(100 * (1 - (L0_poly_sig$count %>% sum / L0_arg_sig_c$count %>% sum)), 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)

```

### 2019-04-04 paper/main.tex
new pipeline is happening, dealing with the data. final numbers and adjusted discussion will have to be done tomorrow

\subsection{L0 polymorphism}

At L0, the type system is mostly aligned with the types of the core language.  We
have observed over 186K monomorphic functions; this accounts for 86\% of our
corpus.  In terms of argument positions, 95\% of arguments are monomorphic.

Table~\ref{tab:L0argcounts} shows the polymorphicity of argument positions
 in R function.  From there, we build up a picture of the polymorphicity of
 functions as a whole, and the data for this function polymorphism can be
 found in Table~\ref{tab:L0funcounts}.

\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
  Monomorphic in Type & 845K & 92.46 \\
  Total Seen & 914K & --- \\
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

\begin{table}[ht]
\label{tab:L0funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\
  \hline
%  Full Monomorphic & 145K & 66.77 \\
  Monomorphic in Type & 187K & 86.4\% \\
%  Monomorphic in Class & 163K & 75.21 \\
%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Function polymorphism in L0.}
\end{table}

@@

\begin{table}[ht]\label{tab:L0toppoly}\centering
\begin{tabular}{lrr}  \hline
type & count & perc \\
\hline
  \D, \sN & 8301 & 17.33 \\
  \D, \l & 6291 & 13.13 \\
  \C, \D & 4338 & 9.06 \\
  \C, \sN & 4162 & 8.69 \\
  \l, \sN & 2912 & 6.08 \\
  \C, \l & 1836 & 3.83 \\
  \C, \I & 1719 & 3.59 \\
  \C, \D, \sN & 1057 & 2.21 \\
  \D, error & 906 & 1.89 \\
  \I, \sN & 797 & 1.66 \\   \hline
\end{tabular}
\caption{Top polymorphic signatures for the L0 type system.}
\end{table}

>>>>

\AT{NOTE: I am updating this with new numbers as we get them.
The fixed tracer is still running, I should be able to fill in final numbers by tomorrow night.}

\subsection{L0}

At L0, the type system is mostly aligned with the types of the core language.  We
have observed over 186K monomorphic functions; this accounts for 86\% of our
corpus.

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
\begin{table}[ht]
\label{tab:L0top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 69806 & 26.3\% \\ 
  \l\ & 55311 & 20.8\% \\ 
  \C & 27789 & 10.5\% \\ 
  \L & 26230 & 9.9\% \\ 
  \sF & 16806 & 6.3\% \\ 
  \sN & 13976 & 5.3\% \\ 
  unevaled & 9125 & 3.4\% \\ 
  \I & 8971 & 3.4\% \\ 
  \sS & 4748 & 1.8\% \\ 
  \D, \I & 3256 & 1.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures.}
\end{table}

We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
Recall that in L0, data frames will appear as lists.
As for ``unevaled'', first recall the discussion that all function arguments are wrapped in promises in R from \AT{where?}.
The ``unevaled'' signature appears whenever an argument was provided but unused, and one might wonder why such a signature might be so common.
As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, so it will appear as ``unevaled''.
The magrittr package's pipe \%$>$\% function is an example of this.
\AT{Move magrittr discussion up.}

One of our chief goals is to generate useful function signatures.
After tracing and capturing unique signatures from function call sites, \AT{recall} that we collapse some of the more obvious ones together.
After this, we are curious to know: {\it how many functions have more than one signature?}
Table~\ref{tab:L0funcounts} has the answer.

\begin{table}[ht]
\label{tab:L0funcounts}
\centering
\begin{tabular}{llr}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 46K & 83\% \\ 
  Total Seen & 56K & --- \\ 
   \hline
\end{tabular}
\caption{Function polymorphism in L0.}
\end{table}

Here, we see that the vast majority of functions are monomorphic in L0.
To choosing how to express the polymorphicity of a function, there are two main options one might consider:
First, having several separate whole-function signatures, e.g. \D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g. (\D $\cup$ \C) $\rightarrow$ \C.
Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument signatures.
To get an idea of how expressing polymorphism at the argument level would look, we ask: {\it what are the most frequently occurring polymorphic argument signatures?}
Table~\ref{tab:L0toppoly} has the answer.

\begin{table}[ht]\label{tab:L0toppoly}\centering
\begin{tabular}{lrr}  \hline
Signature & Count & \% of All Signatures \\
\hline
  \D, \l & 2799 & 1\% \\ 
  \D, \sN & 2290 & 0.9\% \\ 
  \C, \sN & 1527 & 0.6\% \\ 
  \l, \sN & 1193 & 0.4\% \\ 
  \C, \D & 858 & 0.3\% \\ 
  \D, \sF & 413 & 0.2\% \\ 
  \C, \I & 382 & 0.1\% \\ 
  \I, \sN & 378 & 0.1\% \\ 
  \C, \l & 371 & 0.1\% \\ 
  \C, \sF & 334 & 0.1\% \\ 
\end{tabular}
\caption{Top polymorphic argument signatures for the L0 type system.}
\end{table}

Let's unpack some of the signatures in Table~\ref{tab:L0toppoly}:

\begin{itemize}

	\item \D, \l (and \C, \l): L0 doesn't parameterize list types, and it's very likely that this signature describes vectors of doubles and lists of doubles.
	Coercing vectors to lists is trivial in R, and indeed many of the functions with such an argument signature do.
	The story is similar with vectors of characters and lists of characters.

	\item $\square$, \sN: there are a few instances of this ``null'' polymorphism.
	The frequency of these signatures indicates a tendency to create ``optional arguments'', where some argument's default value might be NULL.
	This pattern is quite frequent in R code, \AT{and there are a number of examples sprinkled throughout the paper.}

	\item \C, \D and \C, \I: these are interesting patterns.
	In R, lists can always be indexed with numerics (\L, \D, \I) {\it and also} characters.
	When a list is indexed by character, the character is matched against the element names in the list and the corresponding element is selected.
	If there is no match, the lookup returns NULL.
	Another possible explanation for these signatures are dates, which can be represented by numerics or characters and the value usually has one of the date classes.
	L0 doesn't deal with classes, which are one of the topics of L2.

	\item \D, \sF: this signature appears more strange than it is.
	Some arguments with this signature are either parameters for a function, or the function itself already constructed.
	Another examples, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are parameters for a complex data generation process, or a function to be used during the generation.
	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible.

	\item \C, \sF: another puzzling signature, though its explanation is surprisingly straightforward:
	In R, programmers are free to look functions up {\it by name}, and indeed the frequently encountered {\tt UseMethod} function for dynamic class-based function dispatch is one such example.
	In short, this signature usually represents a function or a function name.

\end{itemize}

Now, we have some understanding of argument polymorphism in L0.
The next question we address is: {\it just how pervasive is argument polymorphism?}.
Refer to Table~\ref{tab:L0argcounts}.

\begin{table}[ht]\label{tab:L0argcounts}\centering
\begin{tabular}{lrl}
  \hline
Signature.Type & Count & Percentage \\ 
  \hline
Monomorphic & 250K & 94\% \\ 
  Total Seen & 266K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

@@
